To be FAIR: Theory Specification Needs an Update

### 2025-03-09 manuscript.tex
submitted version

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 9279}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{longtable, setspace}
\raggedbottom
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={To be FAIR: Theory Specification Needs an Update},
  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Maximilian S. Ernst2,4, Noah N.N. van Dongen5, Felix D. Schönbrodt6, \& Andreas M. Brandmaier2,3,7},
  pdflang={en-EN},
  pdfkeywords={fairtheory, meta theory, theory formation, cumulative science, formal models},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}

@@

This is a preprint paper, generated from Git Commit \# 3cbbaf56. The first and senior author positions were fixed, and remaining author positions randomized using \texttt{set.seed(6999982);\ sample(c("Aaron",\ "Felix",\ "Noah",\ "Maximilian"))}. See the CRediT authorship contributions and GitHub repository for specific author contributions.

@@

\abstract{%
Innovations in open science and meta-science have focused on rigorous \emph{theory testing}, yet methods for specifying, sharing, and iteratively improving theories remain underdeveloped. To address these limitations, we introduce \emph{FAIR theory}: A standard for specifying theories as Findable, Accessible, Interoperable, and Reusable information artifacts. FAIR theories are Findable in well-established archives, Accessible in practical terms (open file formats) and in terms of their ability to be understood, Interoperable for specific purposes, e.g., to guide control variable selection, and Reusable such that they can be iteratively improved through collaborative efforts. This paper adapts the FAIR principles for theory, reflects on the FAIRness of contemporary theoretical practices in psychology, introduces a workflow for FAIRifying theory, and explores FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on the structure and development of theories, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies. We make use of well-established open science infrastructure, including Git for version control, GitHub for collaboration, and Zenodo for archival and search indexing. By applying the principles and infrastructure that have already revolutionized sharing of data and publications to theory, we establish a sustainable, transparent, and collaborative approach to theory development. FAIR theory equips scholars with a standard for systematically specifying and refining theories, bridging a critical gap in open research practices and supporting the renewed interest in theory development in psychology and beyond. FAIR theory provides a structured, cumulative framework for theory development, increasing efficiency and potentially accelerating the pace of cumulative knowledge acquisition.
}

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable (M. D. Wilkinson et al., 2016a) for both humans and computers.
Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software (Lamprecht et al., 2019) and computational workflows (S. R. Wilkinson et al., 2024).
This paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
To this end, we introduce ``FAIR theory'':
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
FAIR theory has the potential to improve the efficiency of scholarly communication and
accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.

@@

The so-called ``replication crisis'' has prompted extensive reforms in psychology and other scientific fields (Lavelle, 2021; Scheel, 2022).
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
These various practices ensure transparent and repeated testing of hypotheses by commiting to an analysis plan in advance.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence (Scheel, Schijen, \& Lakens, 2021).
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
A good example of this is found in ``self-determination theory'' (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures (Deci, 1971).
Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs---autonomy, competence, and relatedness---all still categorized as ``intrinsic'' (Ryan \& Deci, 2000).
The profound difference between these definitions becomes clear when considering the act of changing a child's dirty diaper.
Under the original definition, few caregivers would be considered intrinsically motivated, after all, it's not exactly a joy-filled experience.
Yet, under the expanded definition, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one's child.

Scholars have been raising concerns about the state of theory in psychology for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
One main concern is that theories lack \emph{formalization} (Szollosi \& Donkin, 2021).
When theories are difficult to understand on their own,
without either substantial interpretation or additional background knowledge,
it becomes difficult do derive precise predictions,
and therefore hard to falsify the theory.
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and improvement have been largely overlooked\footnote{Here, we use ``induction'' to refer to inferences from specific observations to general theories. Others have used the term ``abduction'', coined by Peirce to describe ``inference to the best explanation''. For present purposes, however, the terms are interchangeable.}.
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing \emph{FAIR theory} to
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

According to the \emph{empirical cycle} (de Groot, 1961),
a meta-theoretical model of cumulative knowledge acquisition,
research ideally follows a cyclical process with two phases, see Figure \ref{fig:figecs}, panel (a).
In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
In this model, theories are the vehicle of scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.

@@

In a progressive research program (Lakatos, 1971),
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature:
According to Kühberger, Fritz, and Scherndl (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
The remaining 85\% of deductive studies lacked an explicit connection between theory and hypothesis.
Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
in which case it would be straightforward and important to make these explicit (Fried, 2020; Norouzi, Kleinberg, Vermunt, \& Van Lissa, 2024).
Or, perhaps some hypotheses are thoughtlessly reported as part of entrenched research practices (Gigerenzer, Krauss, \& Vitouch, 2004),
but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected (Van Lissa et al., 2020).
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).

Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely connected to, let alone derived from, theory.
Moreover, test results do not routinely contribute to the improvement or rejection of theories.
The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology ``lack the cumulative character of scientific knowledge. They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Meehl, 1978).

@@

The present paper introduces open science methods for theory specification and archival.
Merely publishing theory in a research article does not make it open;
to be open, theory should adhere to established open science standards.
We do so by implementing theories as information artifacts,
archived with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
Metadata are ``data about the data'', they provide information about the nature and content of an information artifact.
Metadata are stored in the repository where the version of record of the FAIR theory is deposited.
FAIR theories are \emph{Findable} via a DOI or by searching the repository they are archived in;
\emph{Accessible} in a machine- and human-readable filetype;
\emph{Interoperable} for specific purposes, for example, within the data analysis environment;
and \emph{Reusable} in the practical and legal sense, so that they may be iteratively improved by the author or by others.
Following the original proposal of Lamprecht and colleagues,
we adapt the FAIR principles for theory, see \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}.
We reflect on the necessary changes (which are minor),
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
and we provide worked examples to encourage their adoption.

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
More rigorous formalization increases theories' \emph{empirical content} (Popper, 2002) because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory (Lewandowsky \& Farrell, 2010).
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Compared to theories expressed in natural language,
formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) ambiguity.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the fidelity and ease with which they are communicated.
Thus, FAIR theory does not require theories to be formal.
At the same time, formal theories are not automatically FAIR.
The FAIR principles pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility.
They can be applied to theories representated in natural language,
as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
Thus, for example, ``grounded theory'', derived from qualitative research,
can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
Conversely, a formal theory is not FAIR if it is confined to a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
FAIR theory is thus consistent with, but does not require, formalization (also see the section on \emph{Accessibility} below).

@@

We propose FAIR theory as an instantiation of modular publishing (Kircz, 1998).
In most fields, the primary unit of scientific communication is the academic paper.
A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
Modular publishing is the practice of making each of these resources available as independent citable \emph{information artifacts} in their own right,
with adequate metadata that is indexed in standardized repositories (Van De Sompel, Payette, Erickson, Lagoze, \& Warner, 2004).
Data sharing is a good example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
Scholars can archive information artifacts in repositories like \href{https://zenodo.org/}{Zenodo},
which was developed by \href{https://home.cern/}{CERN} under the European Union's \href{https://www.openaire.eu/}{OpenAIRE} program (European Organization For Nuclear Research \& OpenAIRE, 2013).
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts - as does, for example, the \href{https://www.crossref.org/}{Crossref} association,
which is used by many academic publishers.
Finally, the DataCite Metadata Schema offers a standard way to document the nature of relationships between information artifacts (DataCite Metadata Working Group, 2024).
For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property \texttt{resourceType:\ dataset},
and cross-reference the published paper with \texttt{relationType:\ IsSupplementTo}.
Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using \texttt{relationType:\ IsDescribedBy}, while the reverse relationship, documented in the canonical reference paper, is \texttt{relationType:\ Describes}. Other types are useful for making relationships between multiple theory objects explicit: If an existing theory is made FAIR without substantial alterations,
the resulting FAIR theory metadata would cross-reference the existing theory as \texttt{relationType:\ IsDerivedFrom}.
If an existing theory is updated, \texttt{relationType:IsNewVersionOf} could be used to reference previous versions.
If a variation of an existing FAIR theory is created, cross-reference it with \texttt{relationType:\ IsVariantFormOf}.

@@

We can take inspiration from the field of computer science for well-established processes for iteratively improving information artifacts.
Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
Git tracks line-by-line changes to text-based files,
and maintains a complete history of those changes.
It has long been argued that Git is particularly well-suited to academic work (Ram, 2013).
Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration (Peikert, Van Lissa, \& Brandmaier, 2021; Van Lissa et al., 2021).
Git provides a useful framework for developing FAIR theory,
because it enables explicitly comparing versions of a file (or: theory),
incorporating changes by different authors,
and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence (Van Lissa, 2023).
Note, however, that the function of Git(Hub) is \emph{not} to archive a FAIR theory.
FAIR-compliant repositories like Zenodo serve the specific purpose of archival.
Thus, while theory development may take place on GitHub, the version of record should be archived in Zenodo, with appropriate metadata.

@@

There have been notable efforts to improve theories' Findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website (Gray, 2017).
Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
Both approaches can be complementary, however.
For example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
Conversely, the database engineering adage ``Lots of Copies Keeps Stuff Safe'' (LOCKSS) implies that it is fine to archive theories in multiple places,
although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.

@@

In sum, authors should make good-faith efforts to make theories as Accessible as possible, in terms of both availability, intelligibility, and freedom from dependencies that cannot be resolved (including dependencies on the author, or manuscripts that can no longer be accessed with reasonable effort).
It is important to recognize that there is an upper bound on interpretability, which means that it is impossible to communicate a theory completely unambiguously.
Nevertheless, scholars should strive to reduce unnecessary ambiguity to the greatest possible extent.
It may benefit scientific discourse to normalize explicit ambiguity (these are things we don't know yet) and anticipate misunderstanding,
to invite others to fill in the blanks and motivate ever further explication of theory.
A theory's Accessibility is increased by
reducing dependencies on (implicit) background knowledge,
explication of assumptions,
formalization,
and explicit cross-references to relevant resources such as papers, ontologies, other related theories, measurement instruments, experimental designs (Lange, Freyer, Musfeld, Schönbrodt, \& Leising, 2025).

@@

If we compare it to the levels of theory formalization (Guest \& Martin, 2021),
it is defined at either the ``theory'' or ``specification'' level.
We can increase the level of formalization, and present an ``implementation'' in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}).
The implementation below describes the model as a directed graph (see also Figure \ref{fig:figecs}a).
Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
and the second half describes their proposed interrelations.
This follows the first two properties of good theory according to Meehl (Meehl, 1990).
We can save this implementation of the empirical cycle to a text file, say \texttt{empirical\_cycle.dot}.

@@

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{fairtheory, meta science, theory formation, open science, formal models\newline\indent Word count: 9211}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{longtable, setspace, subfig}
\raggedbottom
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={To be FAIR: Theory Specification Needs an Update},
  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Maximilian S. Ernst2,4, Noah N.N. van Dongen5, Felix D. Schönbrodt6, \& Andreas M. Brandmaier2,3,7},
  pdflang={en-EN},
  pdfkeywords={fairtheory, meta science, theory formation, open science, formal models},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}

@@

This is a preprint paper, generated from Git Commit \# 985c9433. The first and senior author positions were fixed, and remaining author positions randomized using \texttt{set.seed(6999982);\ sample(c("Aaron",\ "Felix",\ "Noah",\ "Maximilian"))}.

@@

\abstract{%
Innovations in open science and meta-science have focused on rigorous \emph{theory testing}, yet methods for specifying, sharing, and iteratively improving theories remain underdeveloped. To address these limitations, we introduce \emph{FAIR theory}: A standard for specifying theories as Findable, Accessible, Interoperable, and Reusable information artifacts. FAIR theories are Findable in well-established archives, Accessible in practical terms and in terms of their ability to be understood, Interoperable for specific purposes, e.g., to guide control variable selection, and Reusable so that they can be iteratively improved through collaborative efforts. This paper adapts the FAIR principles for theory, reflects on the FAIRness of contemporary theoretical practices in psychology, introduces a workflow for FAIRifying theory, and explores FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on the structure and development of theories, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies. We make use of well-established open science infrastructure, including Git for version control, GitHub for collaboration, and Zenodo for archival and search indexing. By applying the principles and infrastructure that have already revolutionized sharing of data and publications to theory, we establish a sustainable, transparent, and collaborative approach to theory development. FAIR theory equips scholars with a standard for systematically specifying and refining theories, bridging a critical gap in open research practices and supporting the renewed interest in theory development in psychology and beyond. FAIR theory provides a structured, cumulative framework for theory development, increasing efficiency and potentially accelerating the pace of cumulative knowledge acquisition
}

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable (M. D. Wilkinson et al., 2016) for both humans and computers.
Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software (Lamprecht et al., 2019) and computational workflows (S. R. Wilkinson et al., 2024).
This paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
To this end, we introduce ``FAIR theory'':
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
FAIR theory has the potential to improve the efficiency of scholarly communication and
accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.

@@

The so-called ``replication crisis'' has prompted extensive scientific reforms (Lavelle, 2021; Scheel, 2022).
Concern that the abundance of non-replicable findings
was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
These various practices ensure transparent and repeated testing of hypotheses by committing to an analysis plan in advance.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence (Scheel, Schijen, \& Lakens, 2021).
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:

Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so ambiguous that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
A good example of this is found in ``self-determination theory'' (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures (Deci, 1971).
Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs---autonomy, competence, and relatedness---all still categorized as ``intrinsic'' (Ryan \& Deci, 2000).
The difference between these definitions becomes clear when considering the act of changing an infant's dirty diaper.
Under the original definition, one might hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
Under the expanded definition, one might hypothesize the opposite, as the act may fulfill the need for relatedness.
The expanded definition thus enables SDT to absorb potentially falsifying evidence.

Scholars have raised concerns about the state of theory in psychology for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
One main concern is that theories lack \emph{formalization} (Szollosi \& Donkin, 2021).
When theories are ambiguous and hence require either subjective interpretation or additional background knowledge,
it becomes difficult do derive precise predictions,
and therefore hard to falsify the theory.
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
Despite these concerns, scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical processes of theory construction and improvement have been largely overlooked.
The present paper addresses this knowledge gap by applying, for the first time,
open science principles to psychological theory.
We apply introduce \emph{FAIR theory} to
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

According to the \emph{empirical cycle} (de Groot, 1961),
a meta-theoretical model of cumulative knowledge acquisition,
research ideally follows a cyclical process with two phases, see Figure \ref{fig:figecs}a.
In one half of the cycle, labeled the ``Context of Justification'' by Wagenmakers and colleagues, hypotheses derived from theory are tested on data. In the other half of the cycle (the ``Context of Discovery''),
patterns observed in data are generalized to theoretical principles, Figure \ref{fig:figecs}b.
In this model, theories are the vehicle of scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.

@@

In a progressive research program (Lakatos, 1971),
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature:
According to Kühberger, Fritz, and Scherndl (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
The remaining 85\% of deductive studies lacked an explicit connection between theory and hypothesis.
Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
in which case it would be straightforward and important to make these explicit (Fried, 2020; Norouzi, Kleinberg, Vermunt, \& Van Lissa, 2024).
Or, perhaps some hypotheses are reported as part of entrenched research practices (Gigerenzer, Krauss, \& Vitouch, 2004),
but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected (Van Lissa et al., 2020).
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).

Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely connected to, let alone derived from, theory.
Moreover, test results do not routinely contribute to the improvement or rejection of theories.
The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology ``lack the cumulative character of scientific knowledge. They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Meehl, 1978).
The present paper seeks to make theory more tangible and practical by instantiating it as a digital ``object'' that scholars can access, reuse, and update in their daily workflows.

@@

Merely publishing theory in a journal article does not make it open;
to be open, theory should adhere to established open science standards for specification and archival.
We propose to implement theories as information artifacts,
and archive these with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
Metadata are ``data about the data''.
They provide information about the nature and content of an information artifact and are stored in the repository where the version of record of the FAIR theory is deposited.
FAIR theories are \emph{Findable} via a DOI or by searching the repository they are archived in;
\emph{Accessible} in a machine- and human-readable filetype;
\emph{Interoperable} for specific purposes, for example, within the data analysis environment;
and \emph{Reusable} in the practical and legal sense, so that they may be iteratively improved by the author or by others.
Following the original proposal of Lamprecht and colleagues,
we adapt the FAIR principles for theory, see \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}.
We reflect on the necessary changes (which are minor),
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
and we provide worked examples to encourage their adoption.

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
More rigorous formalization increases theories' falsifiability (Popper, 2002) because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory (Lewandowsky \& Farrell, 2010).
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Compared to theories expressed in natural language,
formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) vagueness.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the fidelity and ease with which they are communicated.
Thus, FAIR theory does not require theories to be formal.
At the same time, formal theories are not automatically FAIR.
The FAIR principles
can be applied to theories representated in natural language,
as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
Thus, for example, ``grounded theory'', derived from qualitative research,
can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
Conversely, a formal theory is not FAIR if it is confined to a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
FAIR theory is thus consistent with, but does not require, formalization (also see the section on \emph{Accessibility} below).

@@

We propose FAIR theory as an instantiation of modular publishing (Kircz, 1998).
In most fields, the primary unit of scientific communication is the academic paper.
A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
Modular publishing is the practice of making each of these resources available as independent citable \emph{information artifacts},
with adequate metadata that is indexed in standardized repositories (Van De Sompel, Payette, Erickson, Lagoze, \& Warner, 2004).
Data sharing is an example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
Scholars can archive information artifacts in repositories like \href{https://zenodo.org/}{Zenodo},
which was developed by \href{https://home.cern/}{CERN} under the European Union's \href{https://www.openaire.eu/}{OpenAIRE} program (European Organization For Nuclear Research \& OpenAIRE, 2013).
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts - as does, for example, the \href{https://www.crossref.org/}{Crossref} association,
which is used by many academic publishers.
Finally, the DataCite Metadata Schema offers a standard way to document the nature of relationships between information artifacts (DataCite Metadata Working Group, 2024).
For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property \texttt{resourceType:\ dataset},
and cross-reference the published paper with \texttt{relationType:\ IsSupplementTo}.
Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using \texttt{relationType:\ IsDescribedBy}, while the reverse relationship, documented in the canonical reference paper, is \texttt{relationType:\ Describes}. Other types are useful for making relationships between multiple theory objects explicit: If an existing theory is made FAIR without substantial alterations,
the resulting FAIR theory metadata would cross-reference the existing theory as \texttt{relationType:\ IsDerivedFrom}.
If an existing theory is updated, \texttt{relationType:IsNewVersionOf} could be used to reference previous versions.
If a variation of an existing FAIR theory is created, cross-reference it with \texttt{relationType:\ IsVariantFormOf}.

@@

We can take inspiration from the field of computer science for well-established processes for iteratively improving information artifacts.
Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
Git tracks line-by-line changes to text-based files,
and maintains a complete history of those changes.
It has long been argued that Git is particularly well-suited to academic work (Ram, 2013; Van Lissa et al., 2021).
Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration (Peikert, Van Lissa, \& Brandmaier, 2021; Van Lissa et al., 2021).
Git provides a useful framework for developing FAIR theory,
because it enables explicitly comparing versions of a file (or: theory),
incorporating changes by different authors,
and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence (Van Lissa, 2023).
Note that archival of the version of record is \emph{not} a function of Git(Hub).
While theory development may take place on GitHub, the version of record should be archived in a FAIR-compliant archive like Zenodo, with appropriate metadata.

@@

There have been notable efforts to improve theories' Findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and posted many examples on their website \textless www.theorymaps.org\textgreater{} (Gray, 2017).
Similarly, \href{http://psychomodels.org/}{PsychoModels} seeks to inventorize theories and models in psychology (van Dongen \& Volz, 2025).
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
Both approaches can be complementary, however.
For example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
Conversely, the database engineering adage ``Lots of Copies Keeps Stuff Safe'' (LOCKSS) implies that it is fine to archive theories in multiple places,
although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.

@@

In sum, authors should make good-faith efforts to make theories as Accessible as possible, in terms of both availability, intelligibility, and freedom from dependencies that cannot be resolved (including dependencies on the author, or manuscripts that can no longer be accessed with reasonable effort).
It is important to recognize that there is an upper bound on interpretability, which means that it is impossible to communicate a theory completely unambiguously.
Nevertheless, scholars should strive to reduce unnecessary ambiguity to the greatest possible extent.
It may benefit scientific discourse to normalize explicit ambiguity (these are things we don't know yet) and anticipate misunderstanding,
to invite others to fill in the blanks and motivate ever further explication of theory.
A theory's Accessibility is increased by
reducing dependencies on (implicit) background knowledge,
explication of assumptions,
formalization,
and explicit cross-references to relevant resources such as papers, ontologies, other related theories, measurement instruments, experimental designs (J. Lange, Freyer, Musfeld, Schönbrodt, \& Leising, 2025).

@@

If we compare this to the levels of theory formalization (Guest \& Martin, 2021),
it is defined at either the ``theory'' or ``specification'' level.
We can increase the level of formalization, and present an ``implementation'' in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}).
The implementation below describes the model as a directed graph (see also Figure \ref{fig:figecs}a).
Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
and the second half describes their proposed interrelations.
This follows the first two properties of good theory according to Meehl (Meehl, 1990).
We can save this implementation of the empirical cycle to a text file, say \texttt{empirical\_cycle.dot}.

@@

\bibitem[\citeproctext]{ref-feldEstimatingRelationshipSkill2017}
Feld, J., Sauermann, J., \& de Grip, A. (2017). Estimating the relationship between skill and overconfidence. \emph{Journal of Behavioral and Experimental Economics}, \emph{68}, 18--24. \url{https://doi.org/10.1016/j.socec.2017.03.002}
